Supporting Information:\\ NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2019-04-15 Manuscript/manuscriptPSsuppl.tex
Update manuscriptPSsuppl.tex
Added a section label
Update manuscriptPSsuppl.tex
My edits to S7 on counterions in 5:1 mixtures

\section{Headgroup response to the additional counterions in POPC:POPS (5:1) mixtures}\label{mixtureTOadditionalCIs}
To evaluate counterion binding in different simulation models against experimental data~\cite{roux90},
we plot the headgroup order parameters measured from POPC:POPS 5:1 mixture
as a function of different monovalent ions added to the buffer (Fig. \ref{PSresponseTONaCl}). 
Experimental order parameters of POPC headgroup in the mixture are available as a function
of LiCl and KCl concentrations, while POPS headgroup order parameters are measured also
as a function of NaCl. Lithium interacts more strongly with PS headgroups than other monovalent 
ions~\cite{hauser83,hauser85,roux86,mattai89,roux90}, as also observed for PC headgroups~\cite{cevc90}. 
The different binding behaviour is evident also in the changes of PS headgroup order parameters, which decrease with the addition of lithium 
but increase with the addition of sodium or potassium (Fig. \ref{PSresponseTONaCl}). 
POPC headgroup order parameters exhibit a clear decrease as a function of LiCl concentration
but only modest changes as a function of KCl concentration, indicating singificant 
Li$^+$ binding but only weak K$^+$ binding to the mixture when interpreted using the
electrometer concept~\cite{akutsu81,altenbach84,seelig87}.

In simulations with Berger and CHARMM36 models, the responses of headgroup order parameter
of both POPC and POPS to the added sodium or potassium are more similar to the experiments
with LiCl (Fig. \ref{PSresponseTONaCl}), indicating overestimated binding affinity of sodium and potassium 
in these simulations. The MacRog simulations with potassium exhibit weaker counterion binding affinity
(Fig. \ref{CIdensPSOCmixt}), but significantly larger error bars and
less systematic changes in the order parameters (Fig. \ref{PSresponseTONaCl}).
Similar unsystematic behaviour was also observed in the simulations of Lipid14/17 model
with the additional counterions \cite{POPCpopsLIPID17withKCI,POPCpopsLIPID17withK,POPCpopsLIPID17withNaCI,POPCpopsLIPID17withNa},
for which the data is not shown due to the formation of
ion clusters in water with relatively low (1 M) ion concentrations (Fig. \ref{ionCLUSTERS}).
Appearance of such clusters also in the MagRog simulations with 4 M concentration of KCl
could explain the unsystematic changes of the order parameters in this model with the added KCl.
In conclusion, the results are in line with the section \ref{ciBINDINGsection}
in the main text, suggesting that the MacRog simulations with KCl give the most
realistic surface charge at the lipid bilayer interface among the tested simulation models.
\begin{figure*}[]
  \centering
  \includegraphics[width=17.0cm]{../Figs/CHANGESwithMONVALENTwithPS.eps}
  \caption{\label{PSresponseTONaCl}
    Changes of the PC (left) and PS (right) headgroup order parameters as a function of
    added NaCl, KCl and LiCl from POPC:POPS (5:1) mixture at 298~K
    (except Berger simulations are (4:1) mixture at 310~K).
    The experimental data is from Ref. \citenum{roux90}.
    The values from counterion-only systems are set as a zero point of y-axis.
    To correctly illustrate the significant forking of the $\alpha$-carbon order parameter
    in PS headgroup (bottom, right), the y-axis is tranferred with the same value for both order parameters such that the lower order
    parameter value is at zero.
  }
\end{figure*}
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/CIdensPSOCmixt.eps}
  \caption{  Counterion density distributions from PC:PS mixtures.
\label{CIdensPSOCmixt}
  }
\end{figure}
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/lipid17cluster.png}
  \includegraphics[width=8.0cm]{../Figs/MacRogIONcluster.png}
  \caption{Ion clusters appearing in POPC:POPS (5:1) lipid17/14 simulations with 1 M of NaCl (left)
    and MacRog simulations with 4 M of KCl (right).
\label{ionCLUSTERS}
  }
\end{figure}
\newpage

>>>>

\section{Headgroup response to the additional counterions in POPC:POPS (5:1) mixtures}\label{mixtureTOadditionalCIs}
To evaluate counterion binding in different simulation models against experimental data~\cite{roux90},
we plot the headgroup order parameters measured from POPC:POPS 5:1 mixture
as a function of different monovalent ions added to the buffer (Fig. \ref{PSresponseTONaCl}). 
Experimental order parameters of POPC headgroup in the mixture are available as a function
of LiCl and KCl concentrations, while POPS headgroup order parameters are measured also
in increasing NaCl concentration. Lithium interacts more strongly with PS headgroups than other monovalent 
ions~\cite{hauser83,hauser85,roux86,mattai89,roux90}, as also observed for PC headgroups~\cite{cevc90}. 
The different binding behaviour is evident based on the response of PS headgroup order parameters, which decrease with the addition of lithium 
but increase with the addition of sodium or potassium (Fig. \ref{PSresponseTONaCl}). 
POPC headgroup order parameters exhibit a clear decrease as a function of LiCl concentration
but only modest changes as a function of KCl concentration, indicating singificant 
Li$^+$ binding but only weak K$^+$ binding to the mixture when interpreted using the
electrometer concept~\cite{akutsu81,altenbach84,seelig87}.

In simulations with Berger and CHARMM36 models, the responses of POPC and POPS order parameter
of to added sodium and potassium are not in line with the experiments. Instead, the simulations produces a response similar to experiments conducted in LiCl (Fig. \ref{PSresponseTONaCl}), indicating overestimated binding affinity of sodium and potassium 
in these simulations. The MacRog simulations with potassium exhibit weaker counterion binding affinity
(Fig. \ref{CIdensPSOCmixt}), but significantly larger error bars and
less systematic changes in the order parameters (Fig. \ref{PSresponseTONaCl}).
Similar unsystematic behaviour was also observed in the simulations of Lipid14/17 model
with the additional counterions \cite{POPCpopsLIPID17withKCI,POPCpopsLIPID17withK,POPCpopsLIPID17withNaCI,POPCpopsLIPID17withNa},
for which the data is not shown due to the formation of
ion clusters in water with relatively low (1 M) ion concentrations (Fig. \ref{ionCLUSTERS}).
Appearance of such clusters in the MagRog simulations with 4 M KCl
could explain the unsystematic changes of the order parameters in this model upon increasing KCl.
In conclusion, the results are in line with the section \ref{ciBINDINGsection}
in the main text, suggesting that the MacRog simulations with KCl give the most
realistic surface charge at the lipid bilayer interface among the tested simulation models.
\begin{figure*}[]
  \centering
  \includegraphics[width=17.0cm]{../Figs/CHANGESwithMONVALENTwithPS.eps}
  \caption{\label{PSresponseTONaCl}
    Changes of the PC (left) and PS (right) headgroup order parameters as a function of
    added NaCl, KCl and LiCl from POPC:POPS (5:1) mixture at 298~K
    (except Berger simulations are (4:1) mixture at 310~K).
    The experimental data is from Ref. \citenum{roux90}.
    The values from counterion-only systems are set as a zero point of y-axis.
    To correctly illustrate the significant forking of the $\alpha$-carbon order parameter
    in PS headgroup (bottom, right), the y-axis is shifted with the same value for both order parameters such that the lower order
    parameter value is at zero.
  }
\end{figure*}
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/CIdensPSOCmixt.eps}
  \caption{  Counterion density distributions from PC:PS mixtures.
\label{CIdensPSOCmixt}
  }
\end{figure}
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/lipid17cluster.png}
  \includegraphics[width=8.0cm]{../Figs/MacRogIONcluster.png}
  \caption{Ion clusters appearing in POPC:POPS (5:1) lipid17/14 simulations with 1 M of NaCl (left)
    and MacRog simulations with 4 M of KCl (right).
\label{ionCLUSTERS}
  }
\end{figure}
\newpage
